Abstract

### 2025-10-22 output/Main_text.Rmd
Discard latent space representation supplement
This idea is discarded (for now) for the resubmission.
Apply Grammarly corrections

In this paper, we generalize the multidimensional discrimination (MDISC) and
difficulty (MID) parameters in the Multidimensional 2-Parameter Logistic model
to account for non-identity latent covariances and negatively-keyed items.
We apply Reckase's maximum discrimination point method to define them in an
arbitrary algebraic basis.
Then, we define such basis to be a
geometrical representation of the measured construct.
This results in three different versions of the parameters: The original one,
based on the item parameters solely, one that incorporates the covariance
structure of the latent space, and one that uses the correlation structure
instead. Importantly, we find that the items should be properly represented in
a test space, distinct from the latent space.
We also provide a procedure for the geometrical representation of the items in
the test space, and apply our results to examples from the literature to get a
more accurate representation of the measurement properties of the items.
We recommend using the covariance structure version for describing the
properties of the parameters, and the correlation structure version for
graphical representation.
Finally, we discuss the implications of this generalization for other
multidimensional IRT models and the parallels of our results in common factor
model theory.

@@

for any point, $`r TRAIT_VECTOR`$ or $`r DISCR_VECTOR`$.
This allows representing the coordinates of the latent space vectors
in the test space and, conversely,
the coordinates of the discrimination vectors in the latent space:
By Equations \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
and  \@ref(eq:trait-coords-orth-spaces-eq)
any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
can be expressed in $`r TEST_SPACE_BASIS`$ as

@@

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items with two different latent trait distributions.
(The axes are labelled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
for simplicity, but note that they represent the coordinates of
a latent trait point expressed in test space coordinates,
that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
Their parameters are shown in Table \@ref(tab:example-items-table-out),
both in the original `r MODEL_ACRONYM` metric and
in the multidimensional metric under the two cases,
which differ in the latent trait correlation.
In the first case (columns $`r CORR_ORTH_OUT`$),
the two latent dimensions are independent;
in the second one ($`r CORR_OBL_OUT`$),
the correlation between them is .5.
The same `r MODEL_ACRONYM` parameters are used in both cases,
but their (correlation-based) multidimensional parameters change
with the correlation.
If we used the agnostic version instead, the multidimensional parameters,
given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.

@@

A comparison of panels (a) and (c) allows understanding the effect of
a positive correlation on the $`r MDISC_CORR_PARAM`$
(and consequently the $`r MIL_CORR_PARAM`$) value:
The components of the discrimination parameters tend to
*sum up* in the same direction as they get *more aligned*.
Thus, if the discrimination parameters have the same sign, as in items 1 and 5,
the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
on the contrary, when the discrimination parameters have opposite signs
they tend to cancel each other out and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
as happens with items 2 and 3.
This effect can be especially noticed by comparing
the $`r MDISC_CORR_PARAM_ITEM`$ of items 3 and 5:
Their discrimination parameters are equal in absolute value,
so their $`r MDISC_CORR_PARAM_ITEM`$ values are equal in the orthogonal space;
however, the effect of the correlation shrinks the former and
stretches the latter.
Finally, note that item 4 has one discrimination parameter equal to 0,
so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.

@@

@tezza_modelo_2018 provide us with another example that showcases
the effect of correlated dimensions on negative discrimination parameters.
In their model, dimensions 1 and 4 are estimated to have a correlation of
`r DIMS_CORR`.
With this latent space structure, items that strongly discriminate in either or
both of these two dimensions change drastically their multidimensional
parameters from the agnostic to the covariance-based version
(see Table \@ref(tab:empirical-table-Tezza2018-out))
---this happens,
for example, with items 57 and 60. These items also have different-sign
discrimination parameters in dimensions 1 and 4, so their situation is
similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
Figure \@ref(fig:item-plot-out),
thus the decrease in their $`r MDISC_ITEM`$ parameter.
The effect on the direction is also apparent in the angles these items make with
these dimensions, which change up to almost 20º, compared with at most 7º in the
other two dimensions, and around 1º in most cases.

On the other hand, cases such as items 12 or 59, which have
relatively low discrimination parameters on those two dimensions
(compared with the ones in dimension 2 and 3, respectively) are barely affected.
The most extreme change in these two examples occurs in
the direction of item 12 with respect to dimension 4. We see how the large
discrimination in dimension 1, when combined with the correlation
with dimension 4, induces a decrease in the angular direction
with that dimension.
In this case, because the discrimination in dimension 4 so low, the situation is
very similar to item 4 in Table \@ref(tab:example-items-table-out)
and Figure \@ref(fig:item-plot-out).

@@

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985 and later
extended by @reckase_discriminating_1991
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and
(2) to any case of unchanging monotonicity.
The latter may seem superfluous, given that several examples in the literature
already make use of this generalization
[e.g., the one we have used to illustrate it: @tezza_modelo_2018].
However, this generalization had never been formalized, up to now.
Therefore, we highlight here the interest of showing how
the item psychometric properties are paralleled by their geometrical properties,
and how this holds for any case of unchanging monotonicity
of the item response function.
Regarding the first generalization,
it is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotonicity).
However, they provided no formal proof for this result either.
Our results provide this formal derivation, although only for the specific
case of the `r MODEL_ACRONYM` model.

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameters have the drawback of not fulfilling
the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
namely, the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
This is especially relevant in applications where
potentially different bases are involved
(e.g., multi-group IRT, equating, and linking).
Furthermore, we have found that using this version yields a norm in
the latent space given by the Mahalanobis distance.
This statistical distance not only takes
the latent covariance structure into account,
but makes the latent space norm also invariant to changes of basis.
Given these considerations,
we propose adopting the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., norm invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
However, the representation in the test space allows us to obtain
the direction cosines relative to the original latent trait axes,
which may be useful if they represent actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require transforming the cosines to
the original, non-orthonormal basis afterwards.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024], IRTPRO [@cai_irtpro_2017],
or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution (at least in the exploratory case)
and apply an oblique rotation afterwards.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution by applying
Equation \@ref(eq:dir-cos-vec-cov).
What a test practitioner should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
The latent space, where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix though.
[Note that in the original derivations by @reckase_difficulty_1985;
and @reckase_discriminating_1991, both matrices simplify to the identity matrix,
so there was no need to make this distinction.]
This implies that the axes defined by the test space basis,
and not the ones defined by the latent space basis,
are a geometrical representation of the covariance structure.
Indeed, the cosines of the angles among the test space axes are
the geometric equivalent of the correlations.
This result is both surprising and counterintuitive
---Several instances in the MIRT literature suggest that
"the angle between the axes [of the latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
In fact, we have only defined the latent space as a geometrical artifact
for obtaining the multidimensional parameters.
Interpreting the items in the latent space
could only lead researchers to wrong conclusions.
<!-- #TODO: Iem plots in latent space in Supplement? -->
We hope our results help shed a clearer light on
this often misunderstood topic.

@@

We have seen that taking the latent space structure into account may have a
substantial effect in the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Interpreting the item properties correctly requires a solid understanding of
their relationship with the latent covariance structure,
and we have shown that this can only be achieved
with a proper geometrical representation.
Simple items are not affected much by this issue but,
as MIRT evolves and converges with the common factor psychometric tradition,
we find MIRT models applied more often to new problems,
involving complex multidimensional item structures
(even including items with opposite-signed loadings).
Our results will be especially relevant in those cases,
where the interpretation of the items may be affected more drastically.
For example, the relative importance of an item in scoring a given trait
could be downplayed if it is strongly aligned with that trait dimension
but this alignment is overlooked.
Once again, in multi-group applications,
where differences in the latent space structure is critical,
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Different covariance structures across groups may significantly affect them
due to differences in either the variances or the correlation structure.
Therefore, even when the model parameters are equal across groups,
the multidimensional parameters may not be the same
if the covariance structures differ.
Conversely, the same multidimensional parameters may correspond
to different parameter models.
The subsequent invariance of either the M2PL parameters
or the multidimensional ones may lead to different interpretations of
the same model when applied to different groups.
Although this is an empirical question that requires further research,
it could have significant implications for future theoretical developments
in the understanding of MIRT theory.

>>>>

In this paper, we generalize the multidimensional discrimination (MDISC) and
difficulty (MID) parameters in the Multidimensional 2-Parameter Logistic model
to account for non-identity latent covariances and negatively-keyed items.
We apply Reckase's maximum discrimination point method to define them in an
arbitrary algebraic basis.
Then, we define that basis to be a
geometrical representation of the measured construct.
This results in three different versions of the parameters: The original one,
based on the item parameters solely, one that incorporates the covariance
structure of the latent space, and one that uses the correlation structure
instead. Importantly, we find that the items should be properly represented in
a test space, distinct from the latent space.
We also provide a procedure for the geometrical representation of the items in
the test space, and apply our results to examples from the literature to get a
more accurate representation of the measurement properties of the items.
We recommend using the covariance structure version for describing the
properties of the parameters, and the correlation structure version for
graphical representation.
Finally, we discuss the implications of this generalization for other
multidimensional IRT models and the parallels of our results in common factor
model theory.

@@

for any point, $`r TRAIT_VECTOR`$ or $`r DISCR_VECTOR`$.
This allows representing the coordinates of the latent space vectors
in the test space and, conversely,
the coordinates of the discrimination vectors in the latent space:
By Equations \@ref(eq:change-basis), \@ref(eq:discr-vector-std),
and  \@ref(eq:trait-coords-orth-spaces-eq),
any coordinate vector $`r TRAIT_COORDS`$ in $`r LS_BASIS`$
can be expressed in $`r TEST_SPACE_BASIS`$ as

@@

Up to now, all our derivations have considered $`r N_DIMS`$-dimensional spaces;
that is, multidimensional parameters and item vector coordinates
can be computed in an arbitrary large number of dimensions.
However, as plotting more than two dimensions
is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we limit ourselves to the bidimensional case here.
The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items with two different latent trait distributions.
(The axes are labeled as $`r TRAIT_COMP_1`$ and $`r TRAIT_COMP_2`$
for simplicity, but note that they represent the coordinates of
a latent trait point expressed in test space coordinates,
that is, $`r TRAIT_COORD_1`$ and  $`r TRAIT_COORD_2`$.)
Their parameters are shown in Table \@ref(tab:example-items-table-out),
both in the original `r MODEL_ACRONYM` metric and
in the multidimensional metric under the two cases,
which differ in the latent trait correlation.
In the first case (columns $`r CORR_ORTH_OUT`$),
the two latent dimensions are independent;
in the second one ($`r CORR_OBL_OUT`$),
the correlation between them is .5.
The same `r MODEL_ACRONYM` parameters are used in both cases,
but their (correlation-based) multidimensional parameters change
with the correlation.
If we used the agnostic version instead, the multidimensional parameters,
given by the values under $`r CORR_ORTH_OUT`$, would be the same in both cases.

@@

A comparison of panels (a) and (c) allows understanding the effect of
a positive correlation on the $`r MDISC_CORR_PARAM`$
(and consequently the $`r MIL_CORR_PARAM`$) value:
The components of the discrimination parameters tend to
*sum up* in the same direction as they get *more aligned*.
Thus, if the discrimination parameters have the same sign, as in items 1 and 5,
the $`r MDISC_CORR_PARAM_ITEM`$ value tends to increase;
on the contrary, when the discrimination parameters have opposite signs,
they tend to cancel each other out, and $`r MDISC_CORR_PARAM_ITEM`$ decreases,
as happens with items 2 and 3.
This effect can be especially noticed by comparing
the $`r MDISC_CORR_PARAM_ITEM`$ values of items 3 and 5:
Their discrimination parameters are equal in absolute value,
so their $`r MDISC_CORR_PARAM_ITEM`$ values are equal in the orthogonal space;
however, the effect of the correlation shrinks the former and
stretches the latter.
Finally, note that item 4 has one discrimination parameter equal to 0,
so its $`r MDISC_CORR_PARAM_ITEM`$ is unaffected by the correlation.

@@

@tezza_modelo_2018 provide us with another example that showcases
the effect of correlated dimensions on negative discrimination parameters.
In their model, dimensions 1 and 4 are estimated to have a correlation of
`r DIMS_CORR`.
With this latent space structure, items that strongly discriminate in either or
both of these two dimensions drastically change their multidimensional
parameters from the agnostic to the covariance-based version
(see Table \@ref(tab:empirical-table-Tezza2018-out))
---this happens,
for example, with items 57 and 60. These items also have different-sign
discrimination parameters in dimensions 1 and 4, so their situation is
similar to items 2 and 3 in Table \@ref(tab:example-items-table-out) and
Figure \@ref(fig:item-plot-out),
thus the decrease in their $`r MDISC_ITEM`$ parameter.
The effect on the direction is also apparent in the angles these items make with
these dimensions, which change up to almost 20º, compared with at most 7º in the
other two dimensions, and around 1º in most cases.

On the other hand, cases such as items 12 or 59, which have
relatively low discrimination parameters on those two dimensions
(compared with the ones in dimensions 2 and 3, respectively)
are barely affected.
The most extreme change in these two examples occurs in
the direction of item 12 with respect to dimension 4. We see how the large
discrimination in dimension 1, when combined with the correlation
with dimension 4, induces a decrease in the angular direction
with that dimension.
In this case, because the discrimination in dimension 4 is so low,
the situation is very similar to item 4
in Table \@ref(tab:example-items-table-out) and Figure \@ref(fig:item-plot-out).

@@

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985,
and later extended by @reckase_discriminating_1991,
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and
(2) to any case of unchanging monotonicity.
The latter may seem superfluous, given that several examples in the literature
already make use of this generalization
[e.g., the one we have used to illustrate it: @tezza_modelo_2018].
However, this generalization had never been formalized, up to now.
Therefore, we highlight here the interest of showing how
the item psychometric properties are paralleled by their geometrical properties,
and how this holds for any case of unchanging monotonicity
of the item response function.
Regarding the first generalization,
it is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotonicity).
However, they provided no formal proof for this result either.
Our results provide this formal derivation, although only for the specific
case of the `r MODEL_ACRONYM` model.

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameters have the drawback of not fulfilling
the first of Reckase & McKinley's [-@reckase_discriminating_1991] properties,
namely, the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
This is especially relevant in applications where
potentially different bases are involved
(e.g., multi-group IRT, equating, and linking).
Furthermore, we have found that using this version yields a norm in
the latent space given by the Mahalanobis distance.
This statistical distance not only takes
the latent covariance structure into account,
but also makes the latent space norm invariant to changes of basis.
Given these considerations,
we propose adopting the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, as it has been defined ad hoc to fulfill this specific property,
it lacks the other desirable ones the covariance-based version has
(i.e., norm invariance, and yielding a diagonal latent covariance matrix in
the corresponding orthogonal basis).
Therefore, we recommend its use for graphical representation purposes only,
although in many applications both versions will be coincident, fortunately.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is irrelevant
(e.g., plotting the discrimination parameters for interpretation purposes).

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
Reckase's [-@reckase_difficulty_1985] and
Reckase & McKinley's [-@reckase_discriminating_1991] procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of $`r MDISC_COV_PARAM`$ to a change of basis.
However, the representation in the test space allows us to obtain
the direction cosines relative to the original latent trait axes,
which may be useful if they represent actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require transforming the cosines back to
the original, non-orthonormal basis.
Nevertheless, most available MIRT software solutions,
such as flexMIRT [@cai_flexmirt_2024], IRTPRO [@cai_irtpro_2017],
or R package {mirt} [@chalmers_mirt_2012],
initially estimate an uncorrelated solution (at least in the exploratory case)
and apply an oblique rotation afterward.
Taking advantage of the invariance property,
we may compute the parameters in the orthogonal space
(in fact, this is exactly what {mirt} functions `MDISC()` and `MDIFF()` do).
Then, we may compute the direction cosines in the rotated solution by applying
Equation \@ref(eq:dir-cos-vec-cov).
What a test practitioner should be aware of, though,
is that applying Reckase & McKinley's [-@reckase_discriminating_1991] formula
after any non-orthonormal transformation will yield
a $`r MDISC_AG_PARAM`$ value specific to that basis,
unlike the invariant value of $`r MDISC_COV_PARAM`$.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
The latent space, where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix, though.
[Note that in the original derivations by @reckase_difficulty_1985;
and @reckase_discriminating_1991, both matrices simplify to the identity matrix,
so there was no need to make this distinction.]
This implies that the axes defined by the test space basis,
and not the ones defined by the latent space basis,
are a geometrical representation of the covariance structure.
Indeed, the cosines of the angles among the test space axes are
the geometric equivalent of the correlations.
This result is both surprising and counterintuitive
---Several instances in the MIRT literature suggest that
"the angle between the axes [of the latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
In fact, we have only defined the latent space as a geometrical artifact
for obtaining the multidimensional parameters.
Interpreting the items in the latent space
could only lead researchers to wrong conclusions.
We hope our results help shed a clearer light on
this often misunderstood topic.

@@

We have seen that taking the latent space structure into account may have a
substantial effect on the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Interpreting the item properties correctly requires a solid understanding of
their relationship with the latent covariance structure,
and we have shown that this can only be achieved
with a proper geometrical representation.
Simple items are not affected much by this issue but,
as MIRT evolves and converges with the common factor psychometric tradition,
we find MIRT models applied more often to new problems,
involving complex multidimensional item structures
(even including items with opposite-signed loadings).
Our results will be especially relevant in those cases,
where the interpretation of the items may be affected more drastically.
For example, the relative importance of an item in scoring a given trait
could be downplayed if it is strongly aligned with that trait dimension
but this alignment is overlooked.
Once again, in multi-group applications,
where differences in the latent space structure are critical,
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Different covariance structures across groups may significantly affect them
due to differences in either the variances or the correlation structure.
Therefore, even when the model parameters are equal across groups,
the multidimensional parameters may not be the same
if the covariance structures differ.
Conversely, the same multidimensional parameters may correspond
to different parameter models.
The subsequent invariance of either the M2PL parameters
or the multidimensional ones may lead to different interpretations of
the same model when applied to different groups.
Although this is an empirical question that requires further research,
it could have significant implications for future theoretical developments
in the understanding of MIRT theory.
